---
title: "Lab 05 — Sampling and Bootstrap Distributions"
author: "EAES 480 — Modern Statistics in Earth & Environmental Science"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
editor_options:
  chunk_output_type: inline
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

# Overview

In EAES we almost never observe an entire population. Instead, we take **samples** and compute **point estimates** (like a mean) to infer **population parameters**.

This lab focuses on four core ideas:

1. **Relative error** of point estimates (accuracy)
2. **Standard error (SE)** and how it changes with sample size
3. How **replication** (many repeated samples) creates an approximate **sampling distribution**
4. (Intro) **Bootstrap** distributions: resampling from one sample to approximate uncertainty

## Coding expectations (Week 7+)

This lab uses a mix of:
- **Fill-in-the-blank** (quick checks)
- **Partial pipelines** (you finish the analysis)
- **From-scratch chunks** (you write the full solution)

Use slides + past labs + Discord as needed, but make sure your final code is **your own** and runs **top-to-bottom**.

---

# Data

## Load data (partial pipeline)

Complete the pipeline so `df` is a clean tibble. Filter for 2023, and replace fill values (-9999) with na.

```{r load_data, echo=TRUE, eval=TRUE}
# GOAL: Read the CSV from the repo and clean column names.
# TODO: finish this pipeline.

df <- read_csv("data/us-ams-simple.csv", na = c("-9999")) %>%
  janitor::clean_names() %>%
  dplyr::filter(year_local == 2023)
 

glimpse(df)
```

## Derive time columns 

Write code to add:
- `date` from `year_local` + `doy`
- `month` as labeled abbreviated month (Jan–Dec)
- `day_night` as "Day"/"Night" from `daytime` (0/1)

```{r derive_time, echo=TRUE, eval=FALSE}
# GOAL: Create date/month/day_night columns.
# HINT: DOY is 1-indexed, so use (doy - 1) with origin "YYYY-01-01".

df <- df %>%
 mutate(
    # TODO: create a Date column from year_local and doy
    # HINT: Jan 1 is DOY = 1, so use (doy - 1) with origin = "YYYY-01-01"
    date = as.Date(doy - 1, origin = paste0(year_local, "-01-01")),

    # TODO: create a month column (numeric 1–12 or labeled months)
    month = month(date, label = TRUE, abbr = TRUE),

    # TODO: make a day/night label using daytime (0/1)
    day_night = if_else(daytime == 1, "Day", "Night")
  )

df %>% count(month)
df %>% count(day_night)
```

---

# Population parameters vs point estimates

## 1) Choose variables (fill-in)

Pick **2–3 variables** to analyze. Include at least **one flux** variable (e.g., `nee`, `fc`, `gpp`) and optionally one meteorological variable (e.g., `ta`, `le`).

```{r choose_vars, echo=TRUE, eval=FALSE}
vars <- c(
  "reco",  # e.g., "fc"
  "ta",  # e.g., "ta"
  "fc"   # optional third variable
)

# CHECK: confirm these columns exist (should return character(0))
setdiff(vars, names(df))
```

---

## 2) Compute population parameters (partial pipeline)

Complete the code to compute **population mean** and **population SD** for each selected variable.

Your output should be a tibble named `pop` with columns: `var`, `mean`, `sd`.

```{r pop_params, echo=TRUE, eval=FALSE}
# GOAL: Create a tidy table of population parameters for vars.
# TODO: complete this pipeline.

pop <- df %>%
  summarise(across(
    all_of(vars),
    list(
      mean = ~mean(.x, na.rm = TRUE),
      sd   = ~sd(.x, na.rm = TRUE)
    ),
    .names = "{.col}__{.fn}"
  )) %>%   
  pivot_longer(
    everything(),
    names_to = c("var", "stat"),
    names_sep = "__",
    values_to = "value"
  ) %>%   
  pivot_wider(names_from = stat, values_from = value)       # TODO: pivot_wider

pop
```

**Prompt (2–3 sentences):** Which variable has the largest population SD? What does that imply about the variability of that process?

> The variable with the largest population SD is `ta`, or the air temperature. This implies that air temperature is a relatively variable process, especially when compared to the other variables observed. 

---

# Relative error for point estimates (accuracy)

We will estimate the population mean with a sample mean, then compute **relative error**:

\[
\text{Relative error (\%)} = 100 \times \frac{|\mu - \bar{x}|}{|\mu|}
\]

## 3) One sample: compute relative error 

Write code to:

1. set a seed
2. choose a sample size `n_samp`
3. draw a simple random sample `samp`
4. compute `sample_mean` for each variable in `vars`
5. join to population means from `pop`
6. compute `rel_error_pct` for each variable

Return a tibble with columns: `var`, `sample_mean`, `pop_mean`, `rel_error_pct`.

```{r one_sample_relerr, echo=TRUE, eval=FALSE}
# GOAL: Relative error for point estimates from one random sample.

set.seed(1122)
n_samp <- 100

samp <- df %>%
  slice_sample(n = n_samp)

# get mean point estimate for each variable
est <- samp %>%
  summarise(across(all_of(vars), ~mean(.x, na.rm = T))) %>%
  pivot_longer(everything(), names_to = "var", values_to = "sample_mean")

# compute the relative error for each variable
relerr <- est %>%
  left_join(pop %>% select(var, pop_mean = mean), by = "var") %>%
  mutate(rel_error_pct = 100 * abs(sample_mean - pop_mean) / abs(pop_mean)) %>%
  arrange(desc(rel_error_pct))

relerr
```

**Prompt (3–4 sentences):** Which variable had the largest relative error at your chosen `n_samp`?  
Give a scientific reason why some variables are harder to estimate from a small sample.

> The variable with the largest relative error was `fc`, or the flux carbon in the atmosphere. A scientific reason some variables may be harder to estimate is because there is less predictability and natural change in the atmosphere. For example, something like air temperature normally follows rough trends daily and seasonsally. Something like carbon in the atmosphere though can vary a lot due to climate changes and anthropogenic influnce along with its typical variations.

---

## 4) Relative error vs sample size (partial code + you complete)

We will focus on **one variable** and simulate many repeated studies. The next code chunk is doing quite a lot. Please review the #comments throughout and check you understand generally what each section is doing.

```{r relerr_by_n, echo=TRUE, eval=FALSE}
set.seed(480)

n_grid <- c(25, 100, 300, 1000)   # e.g., c(25, 100, 300, 1000)
reps <- 1000                        # e.g., 500 or 1000

target_var <- "fc"                # e.g., "fch4" or "fc"

# Pull the population vector (remove missing values once, up front)
x_pop <- df[[target_var]]
x_pop <- x_pop[!is.na(x_pop)]

# Population mean for the chosen variable
pop_mean_target <- pop %>%
  filter(var == target_var) %>%
  pull(mean)

# ---- Simulation using explicit for loops ----
sim_list <- vector("list", length(n_grid))

for (i in seq_along(n_grid)) {

  n_samp <- n_grid[i]

  # store relative error from each repeated "study"
  relerrs <- numeric(reps)

  for (j in seq_len(reps)) {

    x_samp <- sample(x_pop, size = n_samp, replace = FALSE)
    xbar <- mean(x_samp)

    relerrs[j] <- 100 * abs(pop_mean_target - xbar) / abs(pop_mean_target)
  }

  sim_list[[i]] <- tibble(
    n = n_samp,
    rel_error_pct = relerrs
  )
}

sim_relerr <- bind_rows(sim_list)

sim_relerr
```

### Summarize + plot (from scratch)

1) Summarize the simulated relative error by `n` using:
- median
- 90th percentile

2) Make a plot that communicates “relative error decreases with n”.

```{r summarize_plot_relerr, echo=TRUE, eval=FALSE}
# GOAL: Summarize and plot the relationship between n and relative error.
# TODO: write your own code.

summary_relerr <- sim_relerr %>%
  group_by(n) %>%
  summarize(
    median_relerr = median(rel_error_pct),
    relerr_90th = quantile(rel_error_pct, 0.9),
    .groups = "drop"
  )
summary_relerr

p_relerr <- 
  ggplot(summary_relerr, aes(x = n)) +
  geom_line(aes(y = median_relerr), color = "deeppink3") +
  geom_point(aes(y = median_relerr), color = "deeppink3") +
  labs(
    x = "Sample Size (n)",
    y = "Relative Error (%)",
    title = "Relative Error vs Sample Size"
  )
p_relerr
```

**Prompt (3–4 sentences):** Describe the pattern. Is improvement from n=25→100 larger than 300→1000?  
Why does accuracy show “diminishing returns” as n increases?

> The pattern is decreasing exponentially, and as size increases, so does the relative error. The improvement of 25→100 to 300→1000 is reflected by this exponential decrease, as when you go from 100 to 1000 samples, the slope becomes much shallower. This plot shows that accuracy is reflective of diminishing returns since as `n` increases and the sample size is relatively representative of the population, further increases in `n` only slightly improve the error value. 

---

# Standard error and sampling distributions

For the sample mean, the standard error is approximately:

\[
SE(\bar{x}) \approx \frac{\sigma}{\sqrt{n}}
\]

where \(\sigma\) is population SD.

## 5) Theory vs simulation (partial pipeline)

Compute:
- theoretical SE
- simulated SD of sample means (over many replications)

```{r se_compare, echo=TRUE, eval=FALSE}
set.seed(480)

# TODO: pick a variable and sample size
target_var <- "fc"
n_samp <- 100
reps <- 1000

pop_sd_target <- pop %>% filter(var == target_var) %>% pull(sd)
pop_mean_target <- pop %>% filter(var == target_var) %>% pull(mean)

means <- replicate(
  reps,
  mean(sample(df[[target_var]], size = n_samp, replace = FALSE), na.rm = TRUE)
)

se_theory <- pop_sd_target / sqrt(n_samp)
se_sim    <- sd(means)

tibble(
  target_var = target_var,
  n = n_samp,
  se_theory = se_theory,
  se_simulated = se_sim
)
```

### Plot the sampling distribution (fill-in)

```{r plot_sampling_dist, echo=TRUE, eval=FALSE}
tibble(mean_est = means) %>%
  ggplot(aes(x = mean_est)) +
  geom_histogram(bins = 20, alpha = 0.85) +
  geom_vline(xintercept = pop_mean_target, linetype = "dashed", linewidth = 1.1) +
  theme_classic(base_size = 18) +
  labs(
    x = paste0("Sample mean of ", target_var),
    y = "Count",
    title = "Sampling distribution of the mean",
    subtitle = paste0("n = ", n_samp, "; reps = ", reps, " (dashed = population mean)")
  )
```

**Prompt (3–5 sentences):** How close were `se_theory` and `se_simulated`?  
What changes if you increase `n_samp` vs increase `reps`?

> `se_theory` and `se_simulated` were very pretty to each other, with a difference of 0.023. When you increase the value of `n_samp`, the values decrease and become more different from each other. If you increase the value of `reps`, the values become more similar to each other, with a difference of only 0.002. 

---

# Bootstrap intuition (one sample, many resamples)

Bootstrap resampling approximates uncertainty **when you only have one sample**.

Key idea: treat your sample as a stand-in for the population and resample **with replacement**.

## 6) Bootstrap distribution of the mean 

Write code to:

- take ONE sample of size `n_samp` from `df`
- generate `B` bootstrap resamples (with replacement) from that sample
- compute bootstrap means
- compute bootstrap SD
- plot the bootstrap distribution

```{r bootstrap_mean, echo=TRUE, eval=FALSE}
set.seed(480)

# TODO: choose settings
target_var <- "fc"
n_samp <- 50
B <- 1000

# one "field campaign" sample
samp <- tibble(x = sample(x_pop, size = n_samp, replace = FALSE))

# bootstrap sampling of the mean
boot_means <- replicate(
  B,
  mean(sample(samp$x, size = n_samp, replace = TRUE))
)

boot_sd <- sd(boot_means)

tibble(
  sample_mean = mean(samp$x),
  boot_sd = boot_sd
)
```

```{r plot_boot, echo=TRUE, eval=FALSE}
# TODO: plot boot_means and mark the sample mean
tibble(boot_means = boot_means) %>%
  ggplot(aes(x = boot_means)) +
  geom_histogram(bins = 20) +
  geom_vline(xintercept = mean(samp$x), linetype = "dashed", linewidth = 1.1)
```

**Prompt (4–6 sentences):** Compare the bootstrap distribution to the sampling distribution from Part III.  
What is the bootstrap trying to approximate? When is bootstrap especially useful in EAES?

> When comparing the two, the peak/median of the bootstrap distribution is a bit over 2, while for the sampling distribution it is only a bit over 0. The bootstrap is trying to approximate the sampling distribution without necessarily knowing the population distribution. It can be useful in EaES because when data sets might only have small sample sizes, it is easier to replicate the samples even with less data. 

---

# Part II — Bootstrap under random vs non-random sampling (SWC)

In Part I, you learned that the bootstrap approximates uncertainty **conditional on the sample you observed**.

Now you will test a key warning:

> Bootstrap can quantify *precision* even when the estimate is *biased*.

We will focus on **soil water content (`swc`)**.

- **Population**: all 2023 half-hour records (after removing `NA`)
- **Sample**: one “field campaign” sample of size `n = 200`
- **Bootstrap**: resample *within the sample* (with replacement) many times

## Preparation (recommended)

```{r part2_prep, echo=TRUE, eval=FALSE}
# GOAL: Ensure df exists (loaded in Part I) and create a clean SWC population vector.

swc_pop <- df$swc
swc_pop <- swc_pop[!is.na(swc_pop)]

pop_mean_swc <- mean(swc_pop)
pop_sd_swc   <- sd(swc_pop)

tibble(
  pop_mean_swc = pop_mean_swc,
  pop_sd_swc   = pop_sd_swc,
  n_pop = length(swc_pop)
)
```

---

## Exercise 1 — Random sample → bootstrap distribution (baseline)

**Task:** Draw a *simple random sample* of `n = 200` from the SWC population, then bootstrap the mean.

```{r ex1_random_bootstrap, echo=TRUE, eval=FALSE}
set.seed(480)

n_samp <- 200
B <- 4000

# 1) Draw a random sample (no replacement) from the population vector
swc_samp_random <- sample(swc_pop, size = n_samp, replace = FALSE)

# 2) Bootstrap the mean (resample WITH replacement from the sample)
boot_means_random <- replicate(
  B,
  mean(sample(swc_samp_random, size = n_samp, replace = TRUE))
)

# 3) Summarize
tibble(
  pop_mean = pop_mean_swc,
  sample_mean = mean(swc_samp_random),
  boot_sd = sd(boot_means_random)
)
```

```{r ex1_plot_random_boot, echo=TRUE, eval=FALSE}
# GOAL: Plot the bootstrap distribution and mark (i) sample mean and (ii) population mean.

tibble(boot_mean = boot_means_random) %>%
  ggplot(aes(x = boot_mean)) +
  geom_histogram(bins = 40, alpha = 0.85) +
  geom_vline(xintercept = mean(swc_samp_random), linetype = "dashed", linewidth = 1.1) +
  geom_vline(xintercept = pop_mean_swc, linetype = "solid", linewidth = 1.1) +
  theme_classic(base_size = 18) +
  labs(
    x = "Bootstrap mean (swc)",
    y = "Count",
    title = "Bootstrap distribution (random sample)",
    subtitle = "Dashed = sample mean; Solid = population mean"
  )
```

**Prompt (3–4 sentences):** Is the bootstrap distribution centered near the population mean?  
What does the bootstrap SD represent here?

> The bootstrap distribution is a little higher than the population mean, but not by a drastic amount. The difference between the two means is a variance of less than one. The standard deviation here represents the possible variance between the bootstrap mean values after sampling. The standard deviation is +/- 0.544, and if the bootstrap mean was around 0.5 lower than it was, it would be much more representative of the population mean. 

---

## Exercise 2 — Non-random sample (summer-only) → bootstrap distribution (bias example)

**Task:** Create a **summer-only** sample (Jun–Aug) of size `n = 200` and bootstrap the mean.

> This simulates a field campaign that only measured during summer.

```{r ex2_summer_sample, echo=TRUE, eval=FALSE}
# GOAL: Build a summer-only pool of SWC observations (Jun–Aug), then sample from that pool.

# TODO: ensure df has a month column.
# If you do not already have `month`, create it from a date column as in Part I.
# HINT: month(date, label = TRUE, abbr = TRUE)


df_summer <- df %>%
  filter(month %in% c("Jun", "Jul", "Aug")) %>%
  filter(!is.na(swc))

swc_summer_pool <- df_summer$swc

length(swc_summer_pool)
```

```{r ex2_summer_bootstrap, echo=TRUE, eval=FALSE}
set.seed(480)

n_samp <- 200
B <- 4000

# 1) Draw a summer-only sample
swc_samp_summer <- sample(swc_summer_pool, size = n_samp, replace = FALSE)

# 2) Bootstrap the mean from that sample
boot_means_summer <- replicate(
  B,
  mean(sample(swc_samp_summer, size = n_samp, replace = TRUE))
)

# 3) Summarize
tibble(
  pop_mean = pop_mean_swc,
  summer_pool_mean = mean(swc_summer_pool),
  sample_mean = mean(swc_samp_summer),
  boot_sd = sd(boot_means_summer)
)
```

```{r ex2_plot_summer_boot, echo=TRUE, eval=FALSE}
# GOAL: Plot summer bootstrap and compare to population mean.

tibble(boot_mean = boot_means_summer) %>%
  ggplot(aes(x = boot_mean)) +
  geom_histogram(bins = 40, alpha = 0.85) +
  geom_vline(xintercept = mean(swc_samp_summer), linetype = "dashed", linewidth = 1.1) +
  geom_vline(xintercept = pop_mean_swc, linetype = "solid", linewidth = 1.1) +
  theme_classic(base_size = 18) +
  labs(
    x = "Bootstrap mean (swc)",
    y = "Count",
    title = "Bootstrap distribution (summer-only sample)",
    subtitle = "Dashed = sample mean; Solid = population mean"
  )
```

**Prompt (4–6 sentences):** Does the bootstrap distribution look “confident” (narrow) even if it’s centered far from the population mean?  
Explain the difference between **precision** and **accuracy** in this example.  
What *kind* of error is summer-only sampling introducing?

> The bootstrap distribution does look confident, but it is much lower than the population mean. The difference between precision and accuracy is that precision is a reference to confidence and variability, which is low here. However, accuracy is a reference to how close the measurement is to the true answer, which is pretty off in this case. The error the summer-only sampling is introducing is selection bias, since only a small part of the total population/year is being sampled. 

---

## Exercise 3 — Stratified sampling (by month) → bootstrap distribution

**Task:** Build a **month-stratified** sample that takes the same number of observations per month.

This is one way to make your sample more representative when the population has strong seasonality.

```{r ex3_stratified_sample, echo=TRUE, eval=FALSE}
set.seed(480)

n_samp <- 240  # divisible by 12 makes this easy (20 per month)
k_per_month <- n_samp / 12

# 1) Create a month-stratified sample from df
df_strat <- df %>%
  filter(!is.na(swc)) %>%
  group_by(month) %>%
  slice_sample(n = k_per_month) %>%
  ungroup()

swc_samp_strat <- df_strat$swc

# CHECK:
df_strat %>% count(month)
```

```{r ex3_strat_bootstrap, echo=TRUE, eval=FALSE}
set.seed(480)

B <- 4000

boot_means_strat <- replicate(
  B,
  mean(sample(swc_samp_strat, size = length(swc_samp_strat), replace = TRUE))
)

tibble(
  pop_mean = pop_mean_swc,
  sample_mean = mean(swc_samp_strat),
  boot_sd = sd(boot_means_strat)
)
```

```{r ex3_plot_strat_boot, echo=TRUE, eval=FALSE}
# GOAL: Plot stratified bootstrap and compare to population mean.

tibble(boot_mean = boot_means_strat) %>%
  ggplot(aes(x = boot_mean)) +
  geom_histogram(bins = 40, alpha = 0.85) +
  geom_vline(xintercept = mean(swc_samp_strat), linetype = "dashed", linewidth = 1.1) +
  geom_vline(xintercept = pop_mean_swc, linetype = "solid", linewidth = 1.1) +
  theme_classic(base_size = 18) +
  labs(
    x = "Bootstrap mean (swc)",
    y = "Count",
    title = "Bootstrap distribution (month-stratified sample)",
    subtitle = "Dashed = sample mean; Solid = population mean"
  )
```

**Prompt (4–6 sentences):** Compare the centers and spreads of the three bootstrap distributions (random vs summer-only vs stratified).  
Which sampling design is most accurate for the annual mean? Which looks most precise?  
What does this teach you about bootstrap and study design?

> I would say the sampling design that is the most accurate for the annual mean is the month-stratified sample. However, the summer-only sample looks the most precise. These results taught me how the results of bootstrapping vary depending on what you're sampling (chosen sample) and the the sample and repitions of a sample can shift results (less and more accurate respectively). In terms of study design, I would say I now have a better understanding on how to produce a relatively accurate model to test data that is representative of a larger population on a small scale. 

---

# Part II wrap-up

**One sentence to remember:**

> Bootstrap measures uncertainty around your estimate, **given your sample** — it cannot rescue a biased sample.


---

# Submission

- Knit your `.Rmd` to HTML.
- Commit and push both the `.Rmd` to your GitHub repo.
